---
title: "Lab 1"
author: "Student names here"
format: pdf
editor: 
  mode: source
---

## Lab Overview

This lab will explore basic `ggplot2` functionality. Please turn in one compiled document (PDF) per group. We will use two datasets related to the recent XXXIII Olympiad (Olympics) held in Paris, France.



## Medal Count Dataset

The first dataset contains medal counts for all countries earning a medal.

```{r, message = F}
library(tidyverse)
medals <- read_csv('https://raw.githubusercontent.com/stat408/Data/main/MedalCount.csv')

```

### 1. (4 points)

Create a figure that tells the story of the medal count at the Paris Olympics.

## Olympic Athlete Dataset

This set of figures will use an Olympic dataset from Kaggle. Additional information is available at <https://www.kaggle.com/datasets/willianoliveiragibin/olympics-2024?resource=download&select=athletes+new.csv>

```{r, message = F}
library(tidyverse)
athletes <- read_csv('https://raw.githubusercontent.com/stat408/Data/main/athletes%20new.csv') |>
  mutate(birth_year = year(birth_date))
```

### 2. (4 points)

Using the `birth_date` variable, create a figure that visualizes the ages of the Olympians. Which sports tend to have the youngest and oldest athletes?


### 3. (4 points)

Create a figure that displays the number of competing athletes from the 12 countries with the most medals. 

```{r}
string_in <- medals$Country[1]

extract_country_abbr <- function(string_in){
  str_split(string_in, '\\(' )[[1]][2] |>
  str_sub(end = -2)
}  

top12 <- sapply(medals$Country[1:12], extract_country_abbr)

top12_athletes <- athletes |>
  filter(country_code %in% top12)

```

Note: I've made this process easier for you by only including athletes from these 12 countries (`r top12`).


## 4. (4 points)

Use the `Q4_data` to visualize the relationship between the number of medals earned by a country against the number of athletes  participating in the Olympics.

```{r}
medals$country_code <- sapply(medals$Country, extract_country_abbr) 

Q4_data <- athletes |> 
  group_by(country_code, country_full) |>
  tally() |>
  rename(num_athletes = n) |>
  left_join(medals, by = 'country_code') |>
  mutate(`Total Medals` = case_when(
    !is.na(`Total Medals`) ~ `Total Medals`,
    is.na(`Total Medals`) ~ 0
  )) |>
  select(country_code, num_athletes, country_full, `Total Medals`) |>
  rename(total_medals = `Total Medals`)

```




## 5. (4 points)

The `athletes` dataset also contains the events the athletes are competing in. See the value for Montana's Katherine Berkoff

`r athletes |> filter(str_detect(name, 'BERKOFF')) |> select(events) |> pull()` 

or gymnast Simone Biles 

`r athletes |> filter(str_detect(name, 'BILES')) |> select(events) |> pull()`

Describe (in words or pseudocode) what you'd need to do and/or what additional information you'd need in order to create a figure that displayed the number of events competed in by athletes that won medals.

